---
title: "Intermediate Elective 2"
author: "Abby Malech"
format: 
  pdf:
    include-in-header: 
       text: |
         \usepackage{fvextra}
         \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
         \DefineVerbatimEnvironment{OutputCode}{Verbatim}{breaklines,commandchars=\\\{\}}
date: "2026-05-15"
execute: 
  warning: false
  messages: false
---

# Set-up

## Read-in packages
```{r}
#| label: load-packages

# general use and cleaning
library(tidyverse)
library(here)
library(janitor)
library(snakecase)

# wrapping axis labels
library(scales)

# reading in .xlsx files
library(readxl)

# visualizing missing data
library(naniar)

# arranging plots
library(patchwork)

```

## Read-in data
```{r}
#| label: load-data

# WEEK 4 data

veg <- read_csv(here("data", "veg.csv"))

metadata <- read_csv(here("data", "vp_veg_metadata.csv"))

# WEEK 5 data
# taxonomic information
taxon_list <- read_csv(here("data", "taxon_list.csv")) 

# aquatic invertebrates
aq_ins <- read_xlsx(here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
                    sheet = "Aquatic Insects") 

# site information
sites <- read_xlsx(here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
                   sheet = "sites")

# water quality information
water_quality <- read_xlsx(here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
                           sheet = "Water Quality",
                           na = c("", "n/a", "N/A", "over", "173+")) 
```

## Class code
```{r}
#| label: class-code

# WEEK 4

# create a new object from veg
pools_explicit0 <- veg |> 
  # filter all occurrences where transect_name contains VP
  filter(str_detect(transect_name, "VP") &
           site == "ncos") |> 
  # complete all possible combinations of year, pool, psoc and distance
  complete(year, transect_name, psoc, transect_distance,
           # fill values of percent cover with 0
           fill = list(percent_cover = 0)
           ) |>
  # group by year, pool, and psoc
  group_by(year, transect_name, psoc) |>
  # calculate total percent cover across all quads
  summarize(sum_pc = sum(percent_cover, na.rm = TRUE)) |>
  # ungroup the data frame
  ungroup() |>
  # create a new column called year_pool from year and transect name
  unite("year_pool", year, transect_name, remove = TRUE) |> 
  # join with metadata data frame
  left_join(metadata, by = "year_pool") |> 
  # calculate mean percent cover from total percent cover/number of quadrats
  mutate(mean_pc = sum_pc/num_quad)

# WEEK 5

# creating new object from sites
ncos_sites <- sites |> 
  # filter to only include NCOS sites sampled four times a year
  filter(sector == "NCOS" & sampling_frequency == "Quarterly")

# creating new clean object from taxon_list
taxon_list_clean <- taxon_list |> 
  # converting taxon name to snake case (no spaces or capital letters,
  # only underscores)
  mutate(verbatim_name = to_any_case(verbatim_name, case = "snake"))

# creating new clean object from water quality
water_quality_clean <- water_quality |> 
  # clean column names
  clean_names() |> 
  # filtering join: only include sites in ncos_sites
  semi_join(ncos_sites, by = "site") |> 
  # create new column to join with later data frames
  unite("date_site", date, site, remove = TRUE) |> 
  # group by date_site column
  group_by(date_site) |> 
  # calculate median pH, salinity, DO
  summarize(med_ph = median(p_h, na.rm = TRUE),
            med_sal = median(salinity_ppt, na.rm = TRUE),
            med_do = median(dissolved_oxygen_mg_l, na.rm = TRUE)) |> 
  # ungroup data frame
  ungroup() |> 
  # filter out salinity outliers
  filter(date_site != "2022-08-12_NVBR") 

# creating new clean object from aquatic inverts
aq_ins_clean <- aq_ins |> 
  # clean column names
  clean_names() |> 
  # filtering join: only include sites occurring in ncos_sites
  semi_join(ncos_sites, by = c("site")) |> 
  # renaming columns
  rename(date = date_on_vial) |> 
  # select columns of interest
  select(date, sample_type, site,
         ostracod,
         copepod,
         hemiptera_corixidae_boatman,
         cladocera,
         nematode,
         diptera_ceratopogonidae, 
         annelida_oligochaete,
         diptera_chironomid,
         annelida_polychaete) |> 
  # filter to only include "filtered beaker" samples
  filter(sample_type %in% c("FB 250", "FB250")) |> 
  # convert the data frame to long format
  pivot_longer(cols = ostracod:annelida_polychaete,
               names_to = "taxon",
               values_to = "abundance") |> 
  # group by date, site, taxon
  group_by(date, site, taxon) |> 
  # calculate average abundance per liter (sum abundance divided by 7.5 liters)
  summarize(ave_lit = sum(abundance, na.rm = TRUE)/7.5) |> 
  # ungroup data frame
  ungroup() |> 
  # create a new column called `date_site` to join with water quality
  unite("date_site", date, site, remove = FALSE) |> 
  # join with water quality
  left_join(water_quality_clean, by = c("date_site")) |> 
  # join with taxon list
  left_join(taxon_list_clean, by = c("taxon" = "verbatim_name")) |> 
  # add full names of sites
  mutate(site_full = case_when(
    site == "NVBR" ~ "North of Venoco Bridge",
    site == "NEC" ~ "South of Venoco Bridge",
    site == "NMC" ~ "Main Channel",
    site == "NPB" ~ "South of Phelps Bridge",
    site == "NPB1" ~ "North of Phelps Bridge",
    site == "NPB2" ~ "Phelps Road",
    site == "NWP" ~  "West Pond",
    site == "NDC" ~ "Devereux Creek"
  )) |> 
  # setting factor levels for site
  mutate(site_full = fct_reorder(site_full, med_sal, .fun = "median", .na_rm = TRUE),
         site_full = fct_rev(site_full))
```

# 1. Explain your inspiration

In 1-2 sentences each, describe:

The visualization I chose for my inspiration was [Tobias Stalder's "Hiking Locations in Washington"](https://github.com/toebR/Tidy-Tuesday/blob/master/hiking/dataviz_hiking.png). This makes sense for the data sets that I am utilizing because I am trying to find the relationship between vegetation and its explanatory variables as well as invertebrate abundance and its explanatory variables. The circular graph will be able to show the abundance of both vegetation and invertebrates, as well as a statistical summary of the mean. 

- which variables from your dataset will be shown in your visualization, and what visual components they map onto (axes, shapes, colors, etc.)


# 2. Plan your figure

Sketch what your figure would look like following your inspiration.

Insert that sketch (as a photo, screenshot, or otherwise) into your document.

# 3. Code your figure

Clean/wrangle your data as needed.

Then, code the figure you decided to create.

Make sure all your code is annotated, and that the output is showing.

At the end of this section, include code to save your final figure as a .jpg or .png file.

4. In 1-3 sentences each, describe:

- your process of using someone else’s code to create a visualization
- challenges you encountered as you made your visualization
- how your visualization differs from visualizations you made in class, for assignments, or for a previous elective



## Vegetation circle visualization

```{r}
#| label: veg


veg_data <- pools_explicit0








```

