---
title: "Intermediate Elective 2"
author: "Abby Malech"
format: 
  pdf:
    include-in-header: 
       text: |
         \usepackage{fvextra}
         \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
         \DefineVerbatimEnvironment{OutputCode}{Verbatim}{breaklines,commandchars=\\\{\}}
date: "2026-05-11"
execute: 
  warning: false
  messages: false
---

# Set-up

## Read-in packages
```{r}
#| label: load-packages



```

## Read-in data
```{r}
#| label: load-data





```

# 1. Explain your inspiration

In 1-2 sentences each, describe:

- which visualization you chose for your inspiration
- why it makes sense for your dataset of choice
- which variables from your dataset will be shown in your visualization, and what visual components they map onto (axes, shapes, colors, etc.)

# 2. Plan your figure

Sketch what your figure would look like following your inspiration.

Insert that sketch (as a photo, screenshot, or otherwise) into your document.

# 3. Code your figure

Clean/wrangle your data as needed.

Then, code the figure you decided to create.

Make sure all your code is annotated, and that the output is showing.

At the end of this section, include code to save your final figure as a .jpg or .png file.

4. In 1-3 sentences each, describe:

- your process of using someone else’s code to create a visualization
- challenges you encountered as you made your visualization
- how your visualization differs from visualizations you made in class, for assignments, or for a previous elective